# **J-PEP Demo: Run the Full Pipeline on Example Data**
vignette: >
  %\VignetteIndexEntry{J-PEP Demo}
  %\VignetteEngine{rmarkdown::html_vignette}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(jpepR)
library(data.table)
library(glue)
library(RSQLite)
library(DBI)
library(igraph)
library(clue)
```

## **Introduction**

The `jpepR` package implements the J-PEP (Joint Pleiotropic and Epigenomic Partitioning) method. It provides tools to:

- Fine-map GWAS traits and store SNP-level posterior inclusion probabilities (PIPs)
- Construct SNP-to-trait ($V_{\text{trait}}$) and SNP-to-tissue ($V_{\text{tissue}}$) matrices
- Run a Bayesian non-negative matrix factorization to decompose SNPs into interpretable biological clusters
- Output matrices of SNP ($W$), trait ($H_{\text{trait}}$), and tissue ($H_{\text{tissue}}$) loadings

The pipeline is modular and includes the following key functions:

| Function                | Description                                                                 |
|-------------------------|-----------------------------------------------------------------------------|
| `finemap_to_sqlite()`   | Fine-maps one trait using P-values and LD blocks                           |
| `make_vtrait()`         | Builds the $V_{\text{trait}}$ matrix from focal + auxiliary traits          |
| `make_vtissue()`        | Builds the $V_{\text{tissue}}$ matrix from epigenomic tracks                |
| `intersect_vmatrices()` | Intersects SNPs shared between $V_{\text{trait}}$ and $V_{\text{tissue}}$   |
| `run_jpep_model()`      | Runs the J-PEP matrix factorization                                          |
| `save_jpep_results()`   | Saves $W$, $H_{\text{trait}}$, and $H_{\text{tissue}}$ matrices to disk     |

## **About the Example Data**

This vignette uses test datasets bundled with the package. These include:

- `focal_sumstats`: summary statistics for one focal trait
- `auxiliary_01` to `auxiliary_05`: summary statistics for five auxiliary traits
- `ld_blocks`: LD blocks to use for fine-mapping
- `exons`: genomic intervals of coding exons, used to exclude exonic SNPs
- `epimap_tracks`: peak files for epigenomic annotations
- `bckg_epi`: epigenomic overlap matrix for background (chr1 SNPs)
- `epimap_metadata` and `epimap_sampleID`: used to map tracks to tissues

These files are sufficient to run the full J-PEP workflow end-to-end.

## **Step 0: Setup and Save Data Locally**

### Load bundled test data
```{r, eval = FALSE}
data(focal_sumstats)
data(auxiliary_01); data(auxiliary_02); data(auxiliary_03)
data(auxiliary_04); data(auxiliary_05)
data(ld_blocks)

data(exons)
data(epimap_tracks)
data(bckg_epi)
data(epimap_metadata)
data(epimap_sampleID)
```

### Set up working directories
```{r, eval = FALSE}
work_dir <- file.path(tempdir(), "jpep_test")
sumstats_dir <- file.path(work_dir, "sumstats")
epi_dir <- file.path(work_dir, "epimap")
output_dir <- file.path(work_dir, "output")

dir.create(sumstats_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(epi_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(output_dir, recursive = TRUE)
```

### Save summary statistics
```{r, eval = FALSE}
fwrite(focal_sumstats, file.path(sumstats_dir, "focal_sumstats.gz"), sep = "\t")
for (i in 1:5) {
  aux_data <- get(glue("auxiliary_0{i}"))
  fwrite(aux_data, file.path(sumstats_dir, glue("auxiliary_0{i}_sumstats.gz")), sep = "\t")
}
```

### Save LD file
```{r, eval = FALSE}
fwrite(ld_blocks, file.path(work_dir, "ld_blocks.txt"), sep = "\t")
```

### Save epigenomic files
```{r, eval = FALSE}
fwrite(exons, file.path(epi_dir, "Homo_sapiens.GRCh37.87_exon.bed"), sep = "\t")
fwrite(epimap_tracks, file.path(epi_dir, "Epimap_tracks.tsv.gz"), sep = "\t")
fwrite(bckg_epi, file.path(epi_dir, "bckg_epi.tsv.gz"), sep = "\t")
fwrite(epimap_metadata, file.path(epi_dir, "Epimap_metadata.txt"), sep = "\t")
fwrite(epimap_sampleID, file.path(epi_dir, "Epimap_sampleID.txt"), sep = "\t")
```

## **Run the Full Pipeline**

### Step 1: Fine-map each trait
This converts GWAS P-values to single-causal posterior inclusion probabilities using block-wise LD structure.
```{r, eval = FALSE}
focal_trait <- "focal"
aux_traits <- paste0("auxiliary_0", 1:5)

for (trait in c(focal_trait, aux_traits)) {
  finemap_to_sqlite(
    trait = trait,
    sumstat_file = file.path(sumstats_dir, paste0(trait, "_sumstats.gz")),
    ld_blocks = file.path(work_dir, "ld_blocks.txt"),
    output_dir = output_dir,
    pip_threshold = 0.01,
    force = TRUE
  )
}
```

### Step 2: Build V_trait
This constructs the SNP-by-trait matrix encoding pleiotropic evidence across traits, aligned to the focal trait.
```{r, eval = FALSE}
make_vtrait(
  focal_trait = focal_trait,
  aux_traits = aux_traits,
  output_dir = output_dir,
  pip_threshold = 0.01,
  aux_pip_thres = 0.5,
  aux_nbr_thres = 1
)
```

### Step 3: Build V_tissue
This generates the SNP-by-tissue matrix by intersecting SNPs with epigenomic peaks and grouping tracks by tissue.
```{r, eval = FALSE}
make_vtissue(
  focal_trait = focal_trait,
  output_dir = output_dir,
  epi_dir = "default",
  pip_threshold = 0.01
)
```

### Step 4: Intersect V_trait and V_tissue
Keep only SNPs that are common to both V_trait and V_tissue matrices.
```{r, eval = FALSE}
vmats <- intersect_vmatrices(
  focal_trait = focal_trait,
  output_dir = output_dir,
  pip_threshold = 0.01
)
```

### Step 5: Run J-PEP
This runs a Bayesian non-negative matrix factorization (bNMF) to discover interpretable clusters of SNPs.
```{r, eval = FALSE}
results <- run_jpep_model(
  V_trait = vmats$V_trait,
  V_tissue = vmats$V_tissue,
  Tolerance = 1e-7,
  k = 15,
  NbrReps = 5,
  Alpha_proj = 10,
  Alpha_cross = 0,
  w_tis = 0.5,
  attempts = 5
)
```

### Step 6: Save Output Matrices
This step writes the $W$, $H_{\text{trait}}$, and $H_{\text{tissue}}$ matrices to disk.
```{r, eval = FALSE}
final_result <- concensus_run(
  runs = results,
  V_trait = vmats$V_trait,
  V_tissue = vmats$V_tissue
)

save_jpep_results(
  result = final_result,
  output_dir = output_dir,
  focal_trait = focal_trait,
  pip_threshold = 0.01
)
```
