---
title: "jpepR-overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{jpepR-overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(jpepR)
```

Introduction

jpepR provides an end-to-end framework for Joint Pleiotropic and Epigenomic Partitioning (J-PEP), a method for identifying biologically interpretable SNP clusters by jointly decomposing pleiotropic and tissue-specific epigenomic signals.

This vignette walks through the full pipeline:

    Fine-map GWAS traits to obtain PIPs

    Construct the pleiotropic matrix V_trait

    Construct the epigenomic matrix V_tissue

    Intersect and align both matrices

    Run the J-PEP decomposition

    Save outputs for downstream interpretation

Setup

You need:

    Summary statistics for the focal and auxiliary traits (in .sumstats.gz format)

    A file defining LD blocks

    Epigenomic annotations (EpiMap BEDs or user-supplied)

    Output directory to save results

focal_trait <- "PASS_T2D_Suzuki_EAS"
aux_traits <- c("WHR_GIANT_2015", "TG_GWAS_2021")
summary_stats_dir <- "/path/to/sumstats"
ld_block_file <- "/path/to/ld_blocks.txt"
output_dir <- "/path/to/output"
epi_dir <- "default"  # or specify a custom directory


1. Fine-mapping: Save PIPs to SQLite

all_traits <- c(focal_trait, aux_traits)
for (trait in all_traits) {
  sumstat_file <- file.path(summary_stats_dir, paste0(trait, "_sumstats.gz"))
  finemap_to_sqlite(trait, sumstat_file, ld_block_file, output_dir)
}

2. Construct V_trait matrix

make_vtrait(
  focal_trait = focal_trait,
  aux_traits = aux_traits,
  output_dir = output_dir
)


3. Construct V_tissue matrix

make_vtissue(
  focal_trait = focal_trait,
  output_dir = output_dir,
  epi_dir = epi_dir
)


4. Align V_trait and V_tissue

vmats <- intersect_vmatrices(
  focal_trait = focal_trait,
  output_dir = output_dir
)


5. Run J-PEP

results <- run_jpep_model(
  V_trait = vmats$V_trait,
  V_tissue = vmats$V_tissue,
  k = 15,
  NbrReps = 10,
  Tolerance = 1e-7,
  Alpha_proj = 10,
  Alpha_cross = 0,
  w_tis = 0.5
)

6. Save Results

save_jpep_results(
  consensus = results,
  focal_trait = focal_trait,
  output_dir = output_dir
)


Output Files

The following outputs will be saved under output_dir/:

    H_traits/: cluster × trait loadings

    H_tissues/: cluster × tissue loadings

    W/: SNP × cluster loadings

You can now use these to interpret disease clusters, annotate SNPs, or compute cross-trait enrichment statistics.
Questions?

For help or bugs, open an issue on the GitHub repository.


---

Once you've pasted this into `vignettes/jpepR-overview.Rmd`, run:

```r
devtools::build_vignettes()

